---
title: "Modeling"
author: "Jamin Goo"
format: html
editor: visual
---

# 1. Introduction

In this section, we will build predictive models to classify whether an individual has diabetes (Diabetes_binary) based on various health indicators.

# 2. Setup and Data Preparation

First, we load the necessary libraries and apply the same data cleaning steps as performed in the EDA phase.

```{r}
library(tidyverse)
library(tidymodels)

# Read the data
diabetes <- read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")

# Apply Data Cleaning and Factor Conversion (Same as EDA)
diabetes_clean <- diabetes |>
  mutate(
    Diabetes_binary = factor(Diabetes_binary, levels = c(0, 1), labels = c("No Diabetes", "Diabetes")),
    HighBP = factor(HighBP, levels = c(0, 1), labels = c("No High BP", "High BP")),
    HighChol = factor(HighChol, levels = c(0, 1), labels = c("No High Chol", "High Chol")),
    CholCheck = factor(CholCheck, levels = c(0, 1), labels = c("No Check", "Check in 5 yrs")),
    Smoker = factor(Smoker, levels = c(0, 1), labels = c("No", "Yes")),
    Stroke = factor(Stroke, levels = c(0, 1), labels = c("No", "Yes")),
    HeartDiseaseorAttack = factor(HeartDiseaseorAttack, levels = c(0, 1), labels = c("No", "Yes")),
    PhysActivity = factor(PhysActivity, levels = c(0, 1), labels = c("No", "Yes")),
    Fruits = factor(Fruits, levels = c(0, 1), labels = c("No", "Yes")),
    Veggies = factor(Veggies, levels = c(0, 1), labels = c("No", "Yes")),
    HvyAlcoholConsump = factor(HvyAlcoholConsump, levels = c(0, 1), labels = c("No", "Yes")),
    AnyHealthcare = factor(AnyHealthcare, levels = c(0, 1), labels = c("No", "Yes")),
    NoDocbcCost = factor(NoDocbcCost, levels = c(0, 1), labels = c("No", "Yes")),
    GenHlth = factor(GenHlth, levels = c(1, 2, 3, 4, 5), 
                     labels = c("Excellent", "Very Good", "Good", "Fair", "Poor")),
    DiffWalk = factor(DiffWalk, levels = c(0, 1), labels = c("No", "Yes")),
    Sex = factor(Sex, levels = c(0, 1), labels = c("Female", "Male")),
    Age = factor(Age),
    Education = factor(Education),
    Income = factor(Income)
  )
```

## 2.1. Data Splitting

We split the data into a training set (70%) and a test set (30%). We set a seed to ensure reproducibility.

```{r}
set.seed(123)

# Stratified split to maintain class balance
data_split <- initial_split(diabetes_clean, prop = 0.70, strata = Diabetes_binary)
train_data <- training(data_split)
test_data  <- testing(data_split)

# Verify the split
nrow(train_data)
nrow(test_data)
```

## 2.2. Cross-Validation Folds

We will use 5-fold cross-validation on the training data to tune our model parameters.

```{r}
cv_folds <- vfold_cv(train_data, v = 5, strata = Diabetes_binary)

# Define the metric set: Log Loss
metrics_eval <- metric_set(mn_log_loss)

```
